A primer to numerical simulations: The perihelion motion of Mercury

### 2017-06-30 tex/MercuryNumerical.tex
Merged versions and included a diff pdf

Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m}\vec F(\vec x)\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
$\Delta t$ by $\Delta t/2$ should not change the result significantly.

@@

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39$ km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in an appropriate unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ {\bf These are earth days, I assume, right?} and distances in $R_0 = 10^{10}$ m.  With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
\begin{equation}
	r_{MS}(0) = 4.6 R_0 \, , \quad
	v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
	\, .
\end{equation}

>>>>

Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} F(\vec x(t))\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
$\Delta t$ by $\Delta t/2$ should not change the result significantly.

@@

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39$ km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in an appropriate unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$, where the time refers to earth days, and distances in $R_0 = 10^{10}$ m.  With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
\begin{equation}
	r_{MS}(0) = 4.6 R_0 \, , \quad
	v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
	\, .
\end{equation}
